Associahedra for finite type cluster algebras \\ and minimal relations between $\b{g}$-vectors

### 2019-06-16 thirdPaper/draft/typeConeAssociahedra.tex
subsection 4.7 sentence
minor changes

% formating the table of contents
\setcounter{tocdepth}{4}
\makeatletter
\def\l@section{\@tocline{1}{5pt}{0pc}{}{}}
\makeatother
\let\oldtocpart=\tocpart
\renewcommand{\tocpart}[2]{\bf\large\oldtocpart{#1}{#2}}
\let\oldtocsection=\tocsection
\renewcommand{\tocsection}[2]{\bf\oldtocsection{#1}{#2}}
\let\oldtocsubsubsection=\tocsubsubsection
\renewcommand{\tocsubsubsection}[2]{\quad\oldtocsubsubsection{#1}{#2}}

@@

\smallskip
\enlargethispage{.2cm}
\item \textbf{Graph associahedra} were defined by M.~Carr and S.~Devadoss~\cite{CarrDevadoss} in connection to C.~De Concini and C.~Procesi's wonderful arrangements~\cite{DeConciniProcesi}. As illustrated in~\cref{fig:specialGraphAssociahedra}, the graph associahedra of certain special families of graphs coincide with well-known families of polytopes: complete graph associahedra are permutahedra, path associahedra are classical associahedra, cycle associahedra are cyclohedra, and star associahedra are stellohedra. The graph associahedra were extended to the nestohedra, which are simple polytopes realizing the nested complex of arbitrary building sets~\cite{Postnikov, FeichtnerSturmfels}. Graph associahedra and nestohedra have been geometrically realized in different ways: by successive truncations of faces of the standard simplex~\cite{CarrDevadoss}, as Minkowski sums of faces of the standard simplex~\cite{Postnikov, FeichtnerSturmfels}, or from their normal fans by exhibiting explicit inequality descriptions~\cite{Devadoss, Zelevinsky}. All these realizations have the same normal fan, called nested fan. We study the type cones of these nested fans (see \cref{subsec:typeConeGA}): we describe their facet-defining inequalities in terms of certain specific pairs of neighboring tubings and count them for the specific families of graph associahedra described earlier. For instance the type cone of the permutahedron coincides with the classical submodular functions on~$2^{[n]}$, described by~$2^{n-2} \binom{n}{2}$ inequalities. In fact, we observe that the type cone of a graphical nested fan is never simplicial except in the case of the classical associahedron (when the graph is the path). Our results describe nevertheless all polytopal realizations of the graphical nested fans.
\end{itemize}

@@

\begin{proof}
Let $\cal{G}$ be a simplicial refinement of $\Fan$. By~\cref{prop:non-simplicial}, the type cone of any coarsening of $\Fan$ is a relatively open cell of the hyperplane arrangement defined by adjacent pairs of~$\cal{G}$. The type cone of the coarsening is not empty whenever it is polytopal. When non-empty, distinct coarsenings give rise to distinct cells (one will have an equality that is a strict inequality in the other), and all lie in the closure of~$\typeCone(\Fan)$ (since they are all in the closure of all the defining half-spaces).
Thus every coarsening describes a relatively open face of~$\ctypeCone(\Fan)$.

@@

Unfortunately, \cref{lem:linearDependencegvectorsCA} is less precise than \cref{prop:exchangeablePairsAsso}. Indeed, which of the two possible linear dependences is satisfied by the $\b{g}$-vectors of~$\cluster \cup \cluster'$ depends on the initial exchange matrix~$\B_\circ$.
However, for cluster algebras of finite type, this linear dependence is independent of the choice of a seed containing~$x$ and~$x'$.
As explained in ~\cref{corollary:UERPforCAproof}, this statement follows from~\cite[Thm.~7.5]{BuanMarshReinekeReitenTodorov}.
We state it here for future reference.
Note that the corresponding statement for cluster variables, instead of~$\b{g}$-vectors, holds by~\cite[Thm.~1.11]{FominZelevinsky-ClusterAlgebrasII}.

@@

Note that the family of brick and $2$-acyclic gentle quivers already contains a lot of relevant examples, including the gentle grid and dissection quivers discussed in \cref{subsubsec:nonkissingComplex}.
In particular, the classical associahedron is the gentle associahedron of a brick and $2$-acyclic gentle quiver.
We will see in~\cref{coro:simplicialTypeConeNKC} that the type cone of the non-kissing fan~$\gvectorFan[\quiver]$ of a brick and $2$-acyclic gentle quiver~$\quiver$ happens to be simplicial, and we will derive in~\cref{coro:allPolytopalRealizationsNKC} a simple description of all polytopal realizations of~$\gvectorFan[\quiver]$.

@@

Conversely, we know from \cref{rem:dimTypeCone} that there are at least~$|\properWalks|-|Q_0|$ extremal exchangeable pairs. Since~$|\strings| = |\distinguishableStrings| = |\properWalks| - |Q_0|$ by~\cref{prop:bijectionStringsWalks}, we conclude that all exchangeable pairs~$\{\cc{\sigma}, \hh{\sigma}\}$ for~$\sigma \in \strings$ are extremal.
\end{proof}

@@

\begin{remark}
\label{exm:nonSimplicialTypeConeNKC}
Note that \cref{prop:interestingExchangeablePairsNKC,prop:extremalExchangeablePairsNKC} and therefore~\cref{coro:simplicialTypeConeNKC,coro:allPolytopalRealizationsNKC} fail when the quiver is not brick or $2$-cyclic.
The smallest exemples are the $2$-cycles whose non-kissing fans are represented in \cref{fig:nonkissingFans2}.
The left one is not brick: it has a non-distinguishable string and a self-kissing walk.
The right one is brick but is $2$-cyclic.

@@

Graph associahedra were defined by M.~Carr and S.~Devadoss~\cite{CarrDevadoss} in connection to C.~De Concini and C.~Procesi's wonderful arrangements~\cite{DeConciniProcesi}.
For a given graph~$\graphG$, the $\graphG$-associahedron~$\Asso[\graphG]$ is a simple polytope whose combinatorial structure encodes the connected subgraphs of~$\graphG$ and their nested structure.
More precisely, the $\graphG$-associahedron is a polytopal realization of the nested complex of~$\graphG$, defined as the simplicial complex of all collections of tubes (connected induced subgraphs) of~$\graphG$ which are pairwise compatible (either nested, or disjoint and non-adjacent).
As illustrated in~\cref{fig:specialGraphAssociahedra}, the graph associahedra of certain special families of graphs coincide with well-known families of polytopes: complete graph associahedra are permutahedra, path associahedra are classical associahedra, cycle associahedra are cyclohedra, and star associahedra are stellohedra.
The graph associahedra were extended to the \defn{nestohedra}, which are simple polytopes realizing the nested complex of arbitrary building sets~\cite{Postnikov, FeichtnerSturmfels}.
Graph associahedra and nestohedra have been geometrically realized in different ways: by successive truncations of faces of the standard simplex~\cite{CarrDevadoss}, as Minkowski sums of faces of the standard simplex~\cite{Postnikov, FeichtnerSturmfels}, or from their normal fans by exhibiting explicit inequality descriptions~\cite{Devadoss, Zelevinsky}.
For a given graph~$\graphG$, the resulting polytopes all have the same normal fan, called nested fan of~$\graphG$: its rays are the characteristic vectors of the tubes, and its cones are generated by characteristic vectors of compatible tubes.
In this section, we describe the type cone of the nested fan of any graph~$\graphG$.

@@

We derive from~\cref{coro:extremalExchangeablePairsGA} the number of extremal exchangeable pairs of the nested fan.
For a tube~$\tube$ of~$\graphG$, we denote 
%by~$\neighbors(\tube)$ the number of neighbors of~$\tube$ and
 by~$\nonDisconnecting(\tube)$ the number of non-disconnecting vertices of~$\tube$.
% In other words, it $\neighbors(\tube)$ (resp.~$\nonDisconnecting(\tube)$) is the number of tubes covering~$\tube$ (resp.~covered by~$\tube$) in the inclusion poset of all tubes of~$\graphG$.
In other words, $\nonDisconnecting(\tube)$ is the number of tubes covered by~$\tube$ in the inclusion poset of all tubes of~$\graphG$.

@@

\begin{lemma}
\label{lemm::mesh mutations and meshes}
The set of positive mesh mutations~$(x,x')$ of \cref{defn::positive mutation} is in bijection with the set~$\meshes$ of~\cref{defn::positive mutation}.
\end{lemma}

@@

\begin{proof}
Let~$T\in\cat$ be a basic cluster tilting object such that~$\CC(T)$ is the initial cluster.
According to~\cref{lemma:positive mutation triangles}, the morphism~$h$ belongs to the ideal~$(\susp T)$.
Therefore, the triangle ${X\to \bigoplus_i E_i \to Y \xrightarrow{h} \susp X}$ induces the relation~$[X]+[Y]=\sum_i [E_i]$ in~$K_0(\cat ; T)$.
\cref{thm::categorification}\,(ii) gives the equality ${\b{g}_{\CC(X)} + \b{g}_{\CC(Y)} = \sum_i\b{g}_{\CC(E_i)}}$, where the~$\b{g}$-vectors are computed with respect to the initial seed~$\CC(T)$.
By \cref{thm::categorification}\,(iii), the middle term of the triangle is uniquely defined, up to isomorphism, by~$X$, and~$Y$, and the triangle is therefore an exchange triangle, with respect to any mutation at~$X$: it does not depend on the choice of a cluster tilting object containing~$X$.
By \cref{thm::categorification}\,(iv) this equality is one of the two equalities in \cref{lem:linearDependencegvectorsCA}.
\end{proof}

@@

As a consequence of \cref{coro::silting simplicial}, the method of \cref{part:geometry} apply and give an explicit description of all realizations of the support~$\tau$-tilting fan of~$\Lambda$.

Another consequence is an algebraic proof of \cref{coro:simplicialTypeConeNKC}, that we restate below.

@@

@@

\addtocontents{toc}{\vspace{.2cm}}
\section*{Acknowledgments}

>>>>

% formating the table of contents
\setcounter{tocdepth}{4}
\makeatletter
\def\l@section{\@tocline{1}{5pt}{0pc}{}{}}
\makeatother
\let\oldtocpart=\tocpart
\renewcommand{\tocpart}[2]{\sc\large\oldtocpart{#1}{#2}}
\let\oldtocsection=\tocsection
\renewcommand{\tocsection}[2]{\bf\oldtocsection{#1}{#2}}
\let\oldtocsubsubsection=\tocsubsubsection
\renewcommand{\tocsubsubsection}[2]{\quad\oldtocsubsubsection{#1}{#2}}

@@

\smallskip
\enlargethispage{.2cm}
\item \textbf{Graph associahedra} were defined by M.~Carr and S.~Devadoss~\cite{CarrDevadoss} in connection to C.~De Concini and C.~Procesi's wonderful arrangements~\cite{DeConciniProcesi}. As illustrated in \cref{fig:specialGraphAssociahedra}, the graph associahedra of certain special families of graphs coincide with well-known families of polytopes: complete graph associahedra are permutahedra, path associahedra are classical associahedra, cycle associahedra are cyclohedra, and star associahedra are stellohedra. The graph associahedra were extended to the nestohedra, which are simple polytopes realizing the nested complex of arbitrary building sets~\cite{Postnikov, FeichtnerSturmfels}. Graph associahedra and nestohedra have been geometrically realized in different ways: by successive truncations of faces of the standard simplex~\cite{CarrDevadoss}, as Minkowski sums of faces of the standard simplex~\cite{Postnikov, FeichtnerSturmfels}, or from their normal fans by exhibiting explicit inequality descriptions~\cite{Devadoss, Zelevinsky}. All these realizations have the same normal fan, called nested fan. We study the type cones of these nested fans (see \cref{subsec:typeConeGA}): we describe their facet-defining inequalities in terms of certain specific pairs of neighboring tubings and count them for the specific families of graph associahedra described earlier. For instance the type cone of the permutahedron coincides with the classical submodular functions on~$2^{[n]}$, described by~$2^{n-2} \binom{n}{2}$ inequalities. In fact, we observe that the type cone of a graphical nested fan is never simplicial except in the case of the classical associahedron (when the graph is the path). Our results describe nevertheless all polytopal realizations of the graphical nested fans.
\end{itemize}

@@

\begin{proof}
Let $\cal{G}$ be a simplicial refinement of $\Fan$. By \cref{prop:non-simplicial}, the type cone of any coarsening of $\Fan$ is a relatively open cell of the hyperplane arrangement defined by adjacent pairs of~$\cal{G}$. The type cone of the coarsening is not empty whenever it is polytopal. When non-empty, distinct coarsenings give rise to distinct cells (one will have an equality that is a strict inequality in the other), and all lie in the closure of~$\typeCone(\Fan)$ (since they are all in the closure of all the defining half-spaces).
Thus every coarsening describes a relatively open face of~$\ctypeCone(\Fan)$.

@@

Unfortunately, \cref{lem:linearDependencegvectorsCA} is less precise than \cref{prop:exchangeablePairsAsso}. Indeed, which of the two possible linear dependences is satisfied by the $\b{g}$-vectors of~$\cluster \cup \cluster'$ depends on the initial exchange matrix~$\B_\circ$.
However, for cluster algebras of finite type, this linear dependence is independent of the choice of a seed containing~$x$ and~$x'$.
As explained in \cref{corollary:UERPforCAproof}, this statement follows from~\cite[Thm.~7.5]{BuanMarshReinekeReitenTodorov}.
We state it here for future reference.
Note that the corresponding statement for cluster variables, instead of~$\b{g}$-vectors, holds by~\cite[Thm.~1.11]{FominZelevinsky-ClusterAlgebrasII}.

@@

Note that the family of brick and $2$-acyclic gentle quivers already contains a lot of relevant examples, including the gentle grid and dissection quivers discussed in \cref{subsubsec:nonkissingComplex}.
In particular, the classical associahedron is the gentle associahedron of a brick and $2$-acyclic gentle quiver.
We will see in \cref{coro:simplicialTypeConeNKC} that the type cone of the non-kissing fan~$\gvectorFan[\quiver]$ of a brick and $2$-acyclic gentle quiver~$\quiver$ happens to be simplicial, and we will derive in \cref{coro:allPolytopalRealizationsNKC} a simple description of all polytopal realizations of~$\gvectorFan[\quiver]$.

@@

Conversely, we know from \cref{rem:dimTypeCone} that there are at least~$|\properWalks|-|Q_0|$ extremal exchangeable pairs. Since~$|\strings| = |\distinguishableStrings| = |\properWalks| - |Q_0|$ by \cref{prop:bijectionStringsWalks}, we conclude that all exchangeable pairs~$\{\cc{\sigma}, \hh{\sigma}\}$ for~$\sigma \in \strings$ are extremal.
\end{proof}

@@

\begin{remark}
\label{exm:nonSimplicialTypeConeNKC}
Note that \cref{prop:interestingExchangeablePairsNKC,prop:extremalExchangeablePairsNKC} and therefore \cref{coro:simplicialTypeConeNKC,coro:allPolytopalRealizationsNKC} fail when the quiver is not brick or $2$-cyclic.
The smallest exemples are the $2$-cycles whose non-kissing fans are represented in \cref{fig:nonkissingFans2}.
The left one is not brick: it has a non-distinguishable string and a self-kissing walk.
The right one is brick but is $2$-cyclic.

@@

Graph associahedra were defined by M.~Carr and S.~Devadoss~\cite{CarrDevadoss} in connection to C.~De Concini and C.~Procesi's wonderful arrangements~\cite{DeConciniProcesi}.
For a given graph~$\graphG$, the $\graphG$-associahedron~$\Asso[\graphG]$ is a simple polytope whose combinatorial structure encodes the connected subgraphs of~$\graphG$ and their nested structure.
More precisely, the $\graphG$-associahedron is a polytopal realization of the nested complex of~$\graphG$, defined as the simplicial complex of all collections of tubes (connected induced subgraphs) of~$\graphG$ which are pairwise compatible (either nested, or disjoint and non-adjacent).
As illustrated in \cref{fig:specialGraphAssociahedra}, the graph associahedra of certain special families of graphs coincide with well-known families of polytopes: complete graph associahedra are permutahedra, path associahedra are classical associahedra, cycle associahedra are cyclohedra, and star associahedra are stellohedra.
The graph associahedra were extended to the \defn{nestohedra}, which are simple polytopes realizing the nested complex of arbitrary building sets~\cite{Postnikov, FeichtnerSturmfels}.
Graph associahedra and nestohedra have been geometrically realized in different ways: by successive truncations of faces of the standard simplex~\cite{CarrDevadoss}, as Minkowski sums of faces of the standard simplex~\cite{Postnikov, FeichtnerSturmfels}, or from their normal fans by exhibiting explicit inequality descriptions~\cite{Devadoss, Zelevinsky}.
For a given graph~$\graphG$, the resulting polytopes all have the same normal fan, called nested fan of~$\graphG$: its rays are the characteristic vectors of the tubes, and its cones are generated by characteristic vectors of compatible tubes.
In this section, we describe the type cone of the nested fan of any graph~$\graphG$.

@@

We derive from \cref{coro:extremalExchangeablePairsGA} the number of extremal exchangeable pairs of the nested fan.
For a tube~$\tube$ of~$\graphG$, we denote 
%by~$\neighbors(\tube)$ the number of neighbors of~$\tube$ and
 by~$\nonDisconnecting(\tube)$ the number of non-disconnecting vertices of~$\tube$.
% In other words, it $\neighbors(\tube)$ (resp.~$\nonDisconnecting(\tube)$) is the number of tubes covering~$\tube$ (resp.~covered by~$\tube$) in the inclusion poset of all tubes of~$\graphG$.
In other words, $\nonDisconnecting(\tube)$ is the number of tubes covered by~$\tube$ in the inclusion poset of all tubes of~$\graphG$.

@@

\begin{lemma}
\label{lemm::mesh mutations and meshes}
The set of positive mesh mutations~$(x,x')$ of \cref{defn::positive mutation} is in bijection with the set~$\meshes$ of \cref{defn::positive mutation}.
\end{lemma}

@@

\begin{proof}
Let~$T\in\cat$ be a basic cluster tilting object such that~$\CC(T)$ is the initial cluster.
According to \cref{lemma:positive mutation triangles}, the morphism~$h$ belongs to the ideal~$(\susp T)$.
Therefore, the triangle ${X\to \bigoplus_i E_i \to Y \xrightarrow{h} \susp X}$ induces the relation~$[X]+[Y]=\sum_i [E_i]$ in~$K_0(\cat ; T)$.
\cref{thm::categorification}\,(ii) gives the equality ${\b{g}_{\CC(X)} + \b{g}_{\CC(Y)} = \sum_i\b{g}_{\CC(E_i)}}$, where the~$\b{g}$-vectors are computed with respect to the initial seed~$\CC(T)$.
By \cref{thm::categorification}\,(iii), the middle term of the triangle is uniquely defined, up to isomorphism, by~$X$, and~$Y$, and the triangle is therefore an exchange triangle, with respect to any mutation at~$X$: it does not depend on the choice of a cluster tilting object containing~$X$.
By \cref{thm::categorification}\,(iv) this equality is one of the two equalities in \cref{lem:linearDependencegvectorsCA}.
\end{proof}

@@

As a consequence of \cref{coro::silting simplicial}, the method of \cref{part:geometry} apply and give an explicit description of all realizations of the support~$\tau$-tilting fan of~$\Lambda$.
Another consequence is an algebraic proof of \cref{coro:simplicialTypeConeNKC}, that we restate below.

@@

In this section, we give two more examples of families of categories to which \cref{theorem::extricats,corollary::meshes positively generate extricats} apply.

@@

\addtocontents{toc}{\vspace{.1cm}}
\section*{Acknowledgments}
